---
title: "cox_subgroup_forest() Tutorial"
author: "Julian Silan"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: kable
    code_folding: show
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

This tutorial walks through `cox_subgroup_forest()` end to end using
`survival::lung`, a built-in dataset of advanced lung cancer patients. It
covers data preparation, forest plot for each covariate and a combined one, 
categorical vs. continuous covariates, custom reference levels, level exclusion,
multiple endpoints, and the main styling arguments.

Source the function first (adjust the path to wherever `forestplot_subgroup_function.R`
lives relative to this `.Rmd` file):

```{r source-function}
source("/Users/julian/Desktop/VHIO/R code/Forestplot/Subgroup/forestplot_subgroup_function.R")
```

```{r libraries}
library(survival)
library(dplyr)
```

# Prepare the data

`survival::lung` needs a little massaging before it's ready for Cox
modelling: `status` is coded 1/2 (censored/dead) rather than 0/1, `sex` is
an unlabelled numeric code, and `ph.ecog` is stored as numeric even though
it is really an ordered categorical performance-status score. We also add a
synthetic treatment arm and a synthetic PFS endpoint so the tutorial can
demonstrate multivariable models and multi-endpoint plots.

```{r prepare-data}
set.seed(42)

lung2 <- survival::lung %>%
  mutate(
    event_os  = status - 1,                       # 1 = death, 0 = censored
    time_os   = time,
    sex       = factor(sex, levels = c(1, 2), labels = c("Male", "Female")),
    ph.ecog   = factor(ph.ecog),                   # 0-3 performance status
    treatment = sample(c("Arm A", "Arm B"), n(), replace = TRUE),
    time_pfs  = pmin(time_os, time_os * runif(n(), 0.4, 0.9)),
    event_pfs = ifelse(time_pfs < time_os, 1, event_os)
  ) %>%
  filter(!is.na(ph.ecog), !is.na(sex), !is.na(age))

#glimpse(lung2)
```

> `time_pfs` / `event_pfs` here are a *simulated* PFS endpoint (PFS event
> always occurs no later than OS) purely so the tutorial has two endpoints
> to work with — `survival::lung` does not contain real PFS data.

# The simplest use

The simplest possible call: one categorical covariate, one endpoint,
univariable model.

```{r sex-only}
res_sex <- cox_subgroup_forest(
  data = lung2,
  covariates = c("sex"),
  labels = c("Sex"),
  treatment = "treatment",
  time = c("time_os"),
  event = c("event_os"),
  endpoint_labels = c("Overall Survival"),
  time_unit = "days",
  palette = c("#0F084B")
)

res_sex$results_table
```

```{r sex-only-plot, fig.height=3}
res_sex$plots$Sex
```


# Several covariates at once (categorical + continuous)


```{r uni-several}
res_sev_cov <- cox_subgroup_forest(
  data = lung2,
  covariates = c("sex", "ph.ecog"),
  labels     = c("Sex", "ECOG PS"),
  treatment = "treatment",
  time = c("time_os"),
  event = c("event_os"),
  endpoint_labels = c("Overall Survival"),
  time_unit = "days",
  palette = c("#0F084B")
)
```


**Table for each covariate**
```{r uni-several-res1}
res_sev_cov$all_result
```

**Table for all covariates together**
```{r uni-several-res2}
res_sev_cov$results_table
```

**Plot for each covariate**
```{r}
res_sev_cov$plots$Sex
res_sev_cov$plots$`ECOG PS`
```

**Plot for all covariates together**
```{r}
res_sev_cov$combined_plots
```

## Specify x-axis

```{r uni-several-axis}
res_x <- cox_subgroup_forest(
  data = lung2,
  covariates = c("sex", "ph.ecog"),
  labels     = c("Sex", "ECOG PS"),
  treatment = "treatment",
  time = c("time_os"),
  event = c("event_os"),
  endpoint_labels = c("Overall Survival"),
  time_unit = "days",
  palette = c("#0F084B"),
  xlim = c(0.25, 4),
  ticks_at = c(0.5, 1, 2)
)

res_x$combined_plots
```

# Excluding a level

`ph.ecog == "3"` has only a handful of patients in `lung`. `exclude_levels`
drops it by setting those values to `NA`, which then falls out through the
existing complete-case filtering used inside each model:

```{r exclude-levels}
res_excl <- cox_subgroup_forest(
  data = lung2,
  covariates = c("sex", "ph.ecog"),
  labels = c("Sex", "ECOG PS"),
  treatment = "treatment",
  time = c("time_os"),
  event = c("event_os"),
  endpoint_labels = c("Overall Survival"),
  time_unit = "days",
  palette = c("#0F084B"),
  exclude_levels = list(ph.ecog = "3"),
  xlim = c(0.25, 4),
  ticks_at = c(0.5, 1, 2),
  length_plot = 60,
  #weights = "weights_ATE", # optional if PS used
  ci_pch = 15, 
  ci_lty = 1,
  ci_lwd = 3, 
  ci_Theight = 0.4, 
  sizes = 1,
  row_height = 1.4,
  covariate_label_row = TRUE
)

res_excl$combined_plots
```

For a **multivariable** model this exclusion drops the affected patients
from the entire model (since multivariable fitting requires complete cases
across every covariate simultaneously) — see the next section.


# Multiple endpoints (OS and PFS) with colour coding

Pass `time` and `event` as vectors, paired by position, along with matching
`endpoint_labels`. With more than one endpoint the plot automatically
colour-codes the confidence intervals per endpoint and adds a legend below
the figure.

```{r two-endpoints, fig.height=8, fig.width=12}
res_two_ep <- cox_subgroup_forest(
  data = lung2,
  covariates = c("sex", "ph.ecog"),
  labels = c("Sex", "ECOG PS"),
  treatment = "treatment",
  time = c("time_os", "time_pfs"),
  event = c("event_os", "event_pfs"),
  endpoint_labels = c("Overall Survival", "Progression-free Survival"),
  time_unit = "days",
  palette = c("#0F084B", "red"),
  #exclude_levels = list(ph.ecog = "3"),
  xlim = c(0.25, 4),
  ticks_at = c(0.5, 1, 2),
  length_plot = 60,
  #weights = "weights_ATE", # optional if PS used
  ci_pch = 15, 
  ci_lty = 1,
  ci_lwd = 3, 
  ci_Theight = 0.4, 
  sizes = 1,
  row_height = 1.4
)

res_two_ep$combined_plots
```


# Styling the plot

A few of the most commonly adjusted arguments:

```{r styled, fig.height=8, fig.width=10}
res_styled <- cox_subgroup_forest(
  data = lung2,
  covariates = c("sex", "ph.ecog"),
  labels = c("Sex", "ECOG PS"),
  treatment = "treatment",
  time = c("time_os", "time_pfs"),
  event = c("event_os", "event_pfs"),
  endpoint_labels = c("Overall Survival", "Progression-free Survival"),
  time_unit = "days",
  palette = c("#0F084B", "red"),
  #exclude_levels = list(ph.ecog = "3"),
  xlim = c(0.25, 4),
  ticks_at = c(0.5, 1, 2),
  #weights = "weights_ATE", # optional if PS used
  ci_pch = 15, 
  ci_lty = 1,
  ci_lwd = 3, 
  ci_Theight = 0.4, 
  sizes = 1,
  row_height = 1.4,
  covariate_label_row = TRUE
)


res_styled$combined_plots
```

- `row_height` — multiplier applied to every body row's height (`1.4` = 40%
  taller); useful when labels wrap or the figure feels cramped.
- `covariate_label_row = TRUE` inserts a dedicated bold row with just the
  covariate name above its levels, instead of bolding the first level row
  in place.
- `xlim` / `ticks_at` let you fix the axis range and tick marks manually
  instead of relying on the automatic (CI-range-based) defaults — handy for
  keeping several related figures on the same scale.


# Adopt length of the forestplot

```{r length, fig.height=8, fig.width=10}
res_length <- cox_subgroup_forest(
  data = lung2,
  covariates = c("sex", "ph.ecog"),
  labels = c("Sex", "ECOG PS"),
  treatment = "treatment",
  time = c("time_os", "time_pfs"),
  event = c("event_os", "event_pfs"),
  endpoint_labels = c("Overall Survival", "Progression-free Survival"),
  time_unit = "days",
  palette = c("#0F084B", "red"),
  #exclude_levels = list(ph.ecog = "3"),
  xlim = c(0.25, 4),
  ticks_at = c(0.5, 1, 2),
  #weights = "weights_ATE", # optional if PS used
  length_plot = 60,
  ci_pch = 15, 
  ci_lty = 1,
  ci_lwd = 3, 
  ci_Theight = 0.4, 
  sizes = 1,
  row_height = 1.4,
  covariate_label_row = TRUE
)
res_length$combined_plots
```


